---
title: "Multivariate Statistical Methods"
subtitle: 'Assignment 4 Canonical correlation analysis'
author: "Ahmet Hakan Akdeve(ahmak554), Jooyoung Lee(joole336), Weng Hang Wong(wonwo535), Zhixuan Duan(zhidu838)"
date: "2019-12-13"
output: pdf_document
---

```{r glb_options, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)
```


## Question: Canonical correlation analysis
### a)
Codes as shown below.

```{r, echo=TRUE}
# import the dataset
bbb <- read.table('/Users/darin/Desktop/multi-var/P10-16.DAT')
bbb <- as.matrix(bbb)

# calculate the corr matrix
aaa <- matrix(0,5,5)

for(i in 1:nrow(bbb))
{
  for(j in 1:ncol(bbb))
  {
    aaa[i,j]<-bbb[i,j]/(sqrt(bbb[i,i])*sqrt(bbb[j,j]))
  }
}

# specific four matrix
s11 <- aaa[1:3,1:3]
s12 <- aaa[1:3,4:5]
s22 <- aaa[4:5,4:5]
s21 <- aaa[4:5,1:3]

# produce the eigenvalues and eigenvectorsof s11
eigen_x <- eigen(s11)

# produce the eigenvalues and eigenvectorsof s22
eigen_y <- eigen(s22)

#  diagonalization
sqrex <- (eigen_x$vectors)%*%diag(eigen_x$values^-.5)%*%solve(eigen_x$vectors) 
sqrey <- (eigen_y$vectors)%*%diag(eigen_y$values^-.5)%*%solve(eigen_y$vectors)

# estimate the eigenvalues and eigenvectors of sqrex and sqrey
m1 <- sqrex%*%s12%*%solve(s22)%*%t(s12)%*%sqrex
m2 <- sqrey%*%t(s12)%*%solve(s11)%*%(s12)%*%sqrey
m1e <- eigen(m1)
m2e <- eigen(m2)


# get a and b
a <- sqrex%*%m1e$vectors
b <- sqrey%*%m2e$vectors

sqrt(0.26764579)
sqrt(0.01575231)
```

We find two non-zero eigenvectors: The highest eigenvalue is 0.268, that is the canonical correlation is 0.518. The second eigenvalue is 0.016, that is the canonical correlation is 0.126.

### b)

```{r, echo=TRUE}
corcoef.test<-function(r, n, p, q, alpha=0.05){
  m<-length(r)
  Q<-rep(0, m)
  lambda <- 1
  
  for (k in m:1){
    lambda<-lambda*(1-r[k]^2); 
    Q[k]<- -log(lambda) 
  }
  
  s<-0
  i<-m
  
  for (k in 1:m){ 
    Q[k]<- (n-k+1-1/2*(p+q+3)+s)*Q[k]
    chi<-1-pchisq(Q[k],(p-k+1)*(q-k+1))
    if (chi>alpha){
      i<-k-1
      break
    }
    s<-s+1/r[k]^2
  }
  return(i)
}

corcoef.test(r=sqrt(m2e$values),n=46,p=3,q=2)
```

From the result, we choose the first pair of canonical correlation.

### c)

The “significant” squared canonical correlation is 0.268 from the result above, which means the correlation between $\ U$ and $\ V$ is 0.268, they groups the variables in such way that the correlation between them is maximized.

### d)

We use the first eigenvectors as weights for the first canonical correlation
($\ X_1$ means glucose intolerance,
$\ X_2$ means insulin response to oral glucose,
$\ X_3$ means insulin resistance,
$\ X_4$ means relative weight,
$\ X_5$ means fasting plasma glucose):

$\ U = 0.436 * X_1 - 0.705 * X_2 + 1.082 * X_3$

$\ V = -1.020 * X_4 + 0.161 * X_5$


### e)

The canonical correlation coefficient cannot explain how much variance each root contribute to the variables, but by checking the significancy of each variable weights in canonical variate, we can assume that this pair of canonical correlated factor is a good summary of the whole dataset.

### f)
There are a few things worth noting, the first is that the correlation between the two sets of variables is reflected by the linear combination of the variables. Second, the Lagrangian equation is used to solve the maximum problem. Third, the canonical correlated variables cannot directly explain the variables.

## Appendix
```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```
